---
title: "world_cup_simulation"
output: 
  html_document
runtime: shiny
---

The World Cup is a tournament in which small differences in team strength and individual match outcomes can have a large impact on the final result. Although it is impossible to predict the winner of every match with certainty, as seen during the previous analysis, simulations can be done to estimate the likelihood of different tournament outcomes.

We simulate the 2026 World Cup using the official group structure and Elo ratings to represent team strength. By repeating the tournament many times through Monte Carlo simulation, we estimate each team's probability of reaching the round of 32, round of 16, quarter-finals, semi-finals and final, as well as winning the title. The aim is not to predict one exact tournament outcome, but to explore a range of plausible outcomes and identify the teams most likely to succeed.

```{r setup_wc_s, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

```{r libraries_wc, include=FALSE}
# load libraries
library(here)
library(gt)
library(dplyr)
library(ggplot2)
library(shiny)
library(plotly)
library(tidyr)

# load data
data <- read.csv(here("data", "wc_elo_ratings.csv"))
```

## Group Composition

We decided to use the same groups as in the 2026 World Cup : the 48 teams are divided into 12 groups, labelled A to L, with four teams in each group. Using the fixed, official groups gives a consistent basis for comparing our simulated outcomes with the tournament's real results.

```{r group_wc}
# teams
teams <- c(
  A1 = "Mexico",
  A2 = "South Africa",
  A3 = "South Korea",
  A4 = "Czechia",
  B1 = "Canada",
  B2 = "Bosnia and Herzegovina",
  B3 = "Qatar",
  B4 = "Switzerland",
  C1 = "Brazil",
  C2 = "Morocco",
  C3 = "Haiti",
  C4 = "Scotland",
  D1 = "United States",
  D2 = "Paraguay",
  D3 = "Australia",
  D4 = "Turkey",
  E1 = "Germany",
  E2 = "Curaçao",
  E3 = "Ivory Coast",
  E4 = "Ecuador",
  F1 = "Netherlands",
  F2 = "Japan",
  F3 = "Sweden",
  F4 = "Tunisia",
  G1 = "Belgium",
  G2 = "Egypt",
  G3 = "Iran",
  G4 = "New Zealand",
  H1 = "Spain",
  H2 = "Cape Verde",
  H3 = "Saudi Arabia",
  H4 = "Uruguay",
  I1 = "France",
  I2 = "Senegal",
  I3 = "Iraq",
  I4 = "Norway",
  J1 = "Argentina",
  J2 = "Algeria",
  J3 = "Austria",
  J4 = "Jordan",
  K1 = "Portugal",
  K2 = "DR Congo",
  K3 = "Uzbekistan",
  K4 = "Colombia",
  L1 = "England",
  L2 = "Croatia",
  L3 = "Ghana",
  L4 = "Panama"
)

# add team's group ID
data$id <- names(teams)[match(data$country, teams)]

# extract group letter
data$letter <- substr(data$id, 1, 1)

# name of the groups
group_letters <- LETTERS[1:12]

# store ratings
rating <- setNames(data$rating, data$country)
```

Each team is assigned an ID consisting of its group letter and a number from 1 to 4. The number represents its position in the group, not its final ranking. The 12 groups are the following :

```{r group_table, warning=FALSE}
# create the text for each group
group_text <- sapply(group_letters, function(letter) {
  group_teams <- teams[paste0(letter, 1:4)]
  
  paste0("**Group ", letter, "**  \n", paste(group_teams, collapse = "  \n"))
})

# arrange the groups into 4x3 matrix
group_table_data <- as.data.frame(matrix(group_text, ncol = 4, byrow = TRUE))

# display the table
groups_table <- group_table_data %>%
  gt() %>%
  fmt_markdown(columns = everything()) %>%
  tab_header(title = "FIFA World Cup 2026 - Groups") %>%
  cols_align(align = "left") %>%
  tab_options(column_labels.hidden = TRUE,
              table.width = pct(100))

# show table
groups_table
```

## Match Model

In the group stage, each team has to play the three other teams in its group. The possible outcomes of a match are a win, a draw or a loss, and teams are ranked according to the points they earn.

To model match outcomes, we first calculate the difference between the two teams' Elo ratings :

$$d = R_A - R_B$$

This rating gap expresses how much stronger (or weaker) team A is than team B. This difference is then used in the Elo formula : $$W_e = \frac{1}{1 + 10^{-d/400}}$$

This formula indicates that for a 400-point rating difference, this corresponds to approximately a 91% expected score for the higher-rated team, which can be interpreted as a very high likelihood of winning. However, it does not separately account for the probability of a draw. We therefore introduce a model for the probability of a draw.

We assume that draws are most likely to occur between teams of similar strength and become less likely as the rating gap increases. We represent this behaviour using a Gaussian function :

$$P(Draw) = D_{max}e^{-d^2/2\sigma^2}$$

Here, $D_{max}$ is the maximum draw probability, reached when the two teams have identical ratings ($d=0$). The parameter $\sigma$ controls how quickly this probability decreases as the rating gap grows in either direction.

Recent World Cups provide a reference for the frequency of draws. Across all matches played, the proportion of draws ranged from approximately 20% to 25% between 2010 and 2022 (*2010: 25%, 2014: 20%, 2018: 20% and 2022: 23%).* We therefore decided to use a maximum draw rate of 30%, slightly above these rates.

To decide on the value of $\sigma$, we compared the resulting draw probabilities across several possible values, as shown in the table below.

```{r sigma, echo=FALSE, warning=FALSE}
# create combinations of Elo differences and sigma values
expand.grid(
  rating_gap = c(0, 100, 200, 300, 400),
  sigma = c(100, 150, 200, 250, 300)
) %>%
  
  # draw probability for each combination
  mutate(
    draw_probability =
      0.30 * exp(-(rating_gap^2) / (2 * sigma^2))
  ) %>%
  
  # reshape 
  pivot_wider(
    names_from = sigma,
    values_from = draw_probability,
    names_prefix = "σ = "
  ) %>%
  
  # create table 
  gt(rowname_col = "rating_gap") %>%
  tab_header(
    title = "Draw probability under different values of σ"
  ) %>%
  tab_stubhead(
    label = "Elo difference"
  ) %>%
  fmt_percent(
    columns = starts_with("σ"),
    decimals = 1
  )%>%
  tab_style(
    style = cell_borders(
      sides = c("left", "right"),
      color = "grey80",
      weight = px(1)
    ),
    locations = cells_body(
      columns = everything()
    )
  ) %>%
  tab_style(
    style = cell_borders(
      sides = c("left", "right"),
      color = "grey70",
      weight = px(1)
    ),
    locations = cells_column_labels(
      columns = everything()
    )
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )%>%
  
  # highlight selected sigma
  tab_style(
    style = cell_fill(color = "#E8F1F5"),
    locations = cells_body(
      columns = `σ = 200`
    )
  ) %>%
  tab_style(
    style = cell_fill(color = "#E8F1F5"),
    locations = cells_column_labels(
      columns = `σ = 200`
    )
  )
```

The table shows how different values of $\sigma$ affect the probability of a draw as the rating gap between two teams increases. Smaller values cause the draw probability to decrease more rapidly, while larger values maintain a higher draw probability. We selected $\sigma = 200$ as a moderate assumption, allowing the draw probability to decrease progressively without declining too quickly or remaining too high for large Elo differences.

Once the draw probability has been determined, we distribute the remaining probability between the two teams using the Elo value :

$$P(A\;wins) = (1-P(draw))W_e \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;P(B\;wins) = (1-P(Draw))(1-W_e)$$

By construction, the three probabilities always sum to 1. We then randomly select a win, draw or loss according to these probabilities to simulate the match outcome.

```{r match_function}
# settings
Dmax <- 0.30
sigma <- 200

# simulate one match
play_match <- function(team_a, team_b) {
  # difference in ratings
  d <- rating[team_a] - rating[team_b]
  
  # probability of a draw
  p_draw <- Dmax * exp (-(d^2) / (2 * sigma^2))
  
  # win expectancy team a
  we_a <- 1 / (1 + 10^(-d / 400))
  
  # win probabilities
  p_a_win <- (1 - p_draw) * we_a
  p_b_win <- (1 - p_draw) * (1 - we_a)
  
  # simulate
  sample(
    c("team_a", "draw", "team_b"),
    size = 1,
    prob = c(p_a_win, p_draw, p_b_win)
  )
}
```

## Group Stage

As mentioned, within each group, every team plays the other three. At the end of the group stage, 32 teams qualify for the knockout phase: the top two teams from each group and the eight best third-placed teams across all groups.

The teams are ranked by points, with three points awarded for a win, one for a draw and none for a loss. In the actual tournament, ties on points are resolved using additional criteria, such as counting goals or the number of yellow and red cards obtained. Since we simulate match outcomes without generating scores or cards, we break ties on points randomly.

```{r group_function}
# simulate all matches in one group
play_group <- function(group_teams) {
  # set points at zero
  points <- setNames(rep(0, 4), group_teams)
  
  # each team plays the other three
  for (i in 1:3) {
    for (j in (i + 1):4) {
      team_a <- group_teams[i]
      team_b <- group_teams[j]
      
      # store result of one match
      result <- play_match(team_a, team_b)
      
      # points
      if (result == "team_a") {
        points[team_a] <- points[team_a] + 3
      }
      
      if (result == "team_b") {
        points[team_b] <- points[team_b] + 3
      }
      
      if (result == "draw") {
        points[team_a] <- points[team_a] + 1
        points[team_b] <- points[team_b] + 1
      }
    }
  }
  
  # rank teams by points and break at random if needed
  ranking <- order(-points, runif(4))
  
  # save ranking
  list(teams = group_teams[ranking], points = points[ranking])
}
```

After simulating the entire group stage, we obtain the 32 qualified teams. Each team receives a new ID combining its group-stage ranking and group letter : 1 for group winners, 2 for runners-up and 3 for qualifying third-placed teams.

```{r group_stage}
# simulate the whole group stage
qualify_r32 <- function() {
  # initialize the qualifiers
  winners <- character(12)
  runners_up <- character(12)
  third_teams <- character(12)
  third_points <- numeric(12)
  
  # play every group
  for (i in 1:12) {
    group_letter <- group_letters[i]
    group_teams <- teams[paste0(group_letter, 1:4)]
    
    # save result
    group_result <- play_group(group_teams)
    winners[i] <- group_result$teams[1]
    runners_up[i] <- group_result$teams[2]
    third_teams[i] <- group_result$teams[3]
    third_points[i] <- group_result$points[3]
  }
  
  # pick the 8 best-third teams
  best_thirds <- order(-third_points, runif(12))[1:8]
  
  # assign new ids
  data.frame(
    new_id = c(
      paste0("1", group_letters),
      paste0("2", group_letters),
      paste0("3", group_letters[best_thirds])
    ),
    team = c(winners, runners_up, third_teams[best_thirds]),
    place = c(rep("winner", 12), rep("runner_up", 12), rep("best_third", 8)),
    stringsAsFactors = FALSE
  )
}
```

## Knockout Stage

Once a team qualifies, it enters the knockout phase, where each match must produce a winner. The winner advances to the next round, while the loser is eliminated. We use the Elo formula directly to model each team's probability of winning the match, as draws are no longer possible.

```{r knockout_function, warning=FALSE}
# simulate one match
knockout_match <- function(team_a, team_b) {
  # rating difference
  d <- rating[team_a] - rating[team_b]
  
  # probability that team a wins
  p_a_win <- 1 / (1 + 10^(-d / 400))
  
  # play match
  if (runif(1) < p_a_win) {
    winner <- team_a
    loser <- team_b
  } else {
    winner <- team_b
    loser <- team_a
  }
  
  # save result
  list(winner = winner, loser = loser)
}
```

The knockout matches follow a predefined schedule. We decided to use the official schedule to maintain consistency with the group composition. Indeed, each team's position in its group determines its possible route through the knockout stage, including which opponents it could face and at which stage.

```{r calendar, warning=FALSE}
# load calendar
calendar_ko <- read.csv(
  here("data", "calendar_ko.csv"),
  stringsAsFactors = FALSE,
  colClasses = "character"
)
```

```{r show_calendar}
# show calendar
htmltools::div(
  style = "max-height: 500px; overflow-y: auto;",
  
  calendar_ko %>%
    gt() %>%
    tab_header(
      title = "World Cup Knockout Stage Calendar"
    ) %>%
    tab_style(
      style = cell_text(weight = "bold"),
      locations = cells_column_labels()
    ) %>%
    tab_style(
      style = cell_borders(
        sides = c("left", "right"),
        color = "grey80",
        weight = px(1)
      ),
      locations = cells_body(
        columns = everything()
      )
    ) %>%
    tab_style(
      style = cell_borders(
        sides = c("left", "right"),
        color = "grey70",
        weight = px(1)
      ),
      locations = cells_column_labels(
        columns = everything()
      )
    )
)
```

The schedule contains each match's number, round and references to team A and team B.

Placing the group winners and runners-up is straightforward, but the eight best third-placed teams are not : their opponents depend on which groups they qualify from, and there are 495 possible combinations. Each third-place slot can only be filled by teams from a specific set of groups, so we use the official allocation table to determine the exact assignment for each combination.

```{r third}
# load third table assignment
third_table <- read.csv(
  here("data", "third_place_assignments_495.csv"),
  colClasses = "character",
  stringsAsFactors = FALSE
)

# show first lines
head(third_table, 5)
```

Each row in the table represents a possible combination of groups from which the 8 best-thirds teams come. For example, the first row applies when these teams come from groups E, F, G, H, I, J, K and L.

```{r thirds_slots}
# assign slots for best thirds
assign_r32_slots <- function(qualifiers) {
  # identify qualifying third-placed teams
  is_third <- qualifiers$place == "best_third"
  third_ids <- qualifiers$new_id[is_third]
  
  # create combination key
  combination <- paste(sort(sub("^3", "", third_ids)), collapse = "")
  
  # find corresponding row in table
  assignment <- third_table[which(third_table$combination == combination), ]
  
  # calendar slots and corresponding columns in table
  slots <- c("3ABCDF",
             "3CDFGH",
             "3CEFHI",
             "3EHIJK",
             "3BEFIJ",
             "3AEHIJ",
             "3EFGIJ",
             "3DEIJL")
  
  columns <- c("vs_1E",
               "vs_1I",
               "vs_1A",
               "vs_1L",
               "vs_1D",
               "vs_1G",
               "vs_1B",
               "vs_1K")
  
  # read third-placed team in each slot
  assigned_ids <- as.character(unlist(assignment[columns], use.names = FALSE))
  
  # keep existing IDs for winners and runners-up
  qualifiers$slot_id <- qualifiers$new_id
  
  # assign calendar slots to third placed teams
  qualifiers$slot_id[is_third] <- slots[match(third_ids, assigned_ids)]
  
  qualifiers
}
```

Once the teams have been assigned their slot IDs for the first knockout round, we simulate the matches in calendar order. The winner of each match advances to the next round and the results are stored to determine the participants in the next matches.

```{r play_knockout}
# play knockout
play_knockout <- function(qualifiers, calendar = calendar_ko) {
  # assign calendar slots to qualified teams
  qualifiers <- assign_r32_slots(qualifiers)
  
  team_by_slot <- setNames(qualifiers$team, qualifiers$slot_id)
  
  # store winners and losers by match ID
  winners <- losers <- setNames(rep(NA_character_, nrow(calendar)), calendar$match_no)
  
  results <- vector("list", nrow(calendar))
  
  # play matches in calendar order
  for (i in seq_len(nrow(calendar))) {
    match_id <- calendar$match_no[i]
    stage <- calendar$round[i]
    
    reference_a <- calendar$team_a[i]
    reference_b <- calendar$team_b[i]
    
    if (stage == "R32") {
      # use the group-stage qualifiers
      team_a <- team_by_slot[reference_a]
      team_b <- team_by_slot[reference_b]
      
    } else if (stage == "Third") {
      # Use the losing semi-finalists
      team_a <- losers[sub("^Runner-up ", "", reference_a)]
      team_b <- losers[sub("^Runner-up ", "", reference_b)]
      
    } else {
      # use the winners of previous matches
      team_a <- winners[reference_a]
      team_b <- winners[reference_b]
    }
    
    team_a <- unname(team_a)
    team_b <- unname(team_b)
    
    # simulate the match
    result <- knockout_match(team_a, team_b)
    
    winners[match_id] <- result$winner
    losers[match_id] <- result$loser
    
    # save the result
    results[[i]] <- data.frame(
      match_no = match_id,
      round = stage,
      team_a = team_a,
      team_b = team_b,
      winner = unname(result$winner),
      loser = unname(result$loser),
      stringsAsFactors = FALSE
    )
  }
  
  do.call(rbind, results)
}
```

## Simulation

Finally, we run the Monte Carlo simulation, repeating the tournament many times to estimate each team's probability of reaching each stage and winning the title. For each simulation, we record which teams reach each stage and which teams win the tournament. After all simulations, we estimate each probability by dividing the number of times a team reaches that stage by the total number of simulations.

```{r monte_carlo_function}
# seed not fixed to allow multiple simulations
set.seed(NULL)

# monte carlo function
run_cup_mc <- function(number_of_simulations) {
  
  # stages we want probabilities
  stages <- c("R32", "R16", "QF", "SF", "Final", "Champion")
  
  # track each team's progress
  results <- matrix(
    0,
    nrow = length(teams),
    ncol = length(stages),
    dimnames = list(unname(teams), stages)
  )
  
  # repeat complete tournament
  for (i in seq_len(number_of_simulations)) {
    
    # simulate group stage
    qualified_teams <- qualify_r32()
    
    # simulate knockout stage
    knockout_results <- play_knockout(
      qualifiers = qualified_teams,
      calendar = calendar_ko
    )
    
    # collect teams that reached each stage in this tournament
    reached <- list(
      R32 = qualified_teams$team,
      
      R16 = knockout_results$winner[
        knockout_results$round == "R32"
      ],
      
      QF = knockout_results$winner[
        knockout_results$round == "R16"
      ],
      
      SF = knockout_results$winner[
        knockout_results$round == "Quarter"
      ],
      
      Final = knockout_results$winner[
        knockout_results$round == "Semi"
      ],
      
      Champion = knockout_results$winner[
        knockout_results$round == "Final"
      ]
    )
    
    
    # update counters for each stage
    for (stage in stages) {
      participants <- reached[[stage]]
      results[participants, stage] <-
        results[participants, stage] + 1
    }
  }
  
  # convert counts in percentages
  percentages <- 100 * results / number_of_simulations
  
  # results table
  output <- data.frame(
    team = rownames(percentages),
    elo = unname(rating[rownames(percentages)]),
    percentages,
    row.names = NULL,
    stringsAsFactors = FALSE
  )
  
  # sort before rounding
  output <- output[
    order(-output$Champion, -output$Final),
  ]
  
  # apply rounding
  output[stages] <- lapply(output[stages], round, digits = 1)
  rownames(output) <- NULL
  
  output
}
```

## Interactive App

The following interactive application allows users to choose the number of simulations and a random seed. Using the same seed and settings reproduces the predictions, while changing the seed generates a different simulation run.

The first tab, "Tournament predictions", displays the favourites' estimated title probabilities. The second tab, "Explore a team", lets users select a team and view its probability of reaching each stage.

```{r simulator-ui, echo=FALSE}
# Shiny user interface
tagList(
  
  # Styling for the application
  tags$style(
    HTML(
      "
      .wc-app {
        width: 100%;
        height: 700px;
        overflow-y: auto;
        overflow-x: hidden;
        padding: 15px;
        border: 1px solid #dfe5eb;
        border-radius: 12px;
        background: white;
        box-sizing: border-box;
      }
      "
    )
  ),
  
  # Main application container
  div(
    class = "wc-app",
    
    sidebarLayout(
      
      # Simulation settings
      sidebarPanel(
        
        numericInput(
          "n_simulations",
          "Number of simulations",
          value = 1000,
          min = 100,
          max = 50000,
          step = 100
        ),
        
        numericInput(
          "simulation_seed",
          "Random seed",
          value = 123,
          min = 1,
          max = 1000000,
          step = 1
        ),
        
        actionButton(
          "run_mc",
          "Run simulations",
          class = "btn-primary"
        ),
        
        helpText(
          "The same seed and settings reproduce the same predictions."
        ),
        
        width = 3
      ),
      
      # Simulation results
      mainPanel(
        
        tabsetPanel(
          
          # TAB 1: Tournament predictions
          tabPanel(
            "Tournament predictions",
            
            br(),
            
            textOutput("simulation_status"),
            
            plotlyOutput(
              "favourites_chart",
              height = "500px"
            )
          ),
          
          # TAB 2: Explore an individual team
          tabPanel(
            "Explore a team",
            
            br(),
            
            selectInput(
              "selected_team",
              "Select a team",
              choices = sort(unname(teams)),
              selected = "Switzerland"
            ),
            
            h3(
              textOutput("team_title")
            ),
            
            textOutput("team_details"),
            
            plotlyOutput(
              "team_chart",
              height = "400px"
            )
          )
        ),
        
        width = 9
      )
    )
  )
)
```

```{r simulator, echo=FALSE}
run_mc_shiny <- function(n) {
  # Stages tracked for every team
  stages <- c("R32", "R16", "QF", "SF", "Final", "Champion")
  
  # Start all counters at zero
  counts <- matrix(
    0,
    nrow = length(teams),
    ncol = length(stages),
    dimnames = list(unname(teams), stages)
  )
  
  # Update the progress bar approximately 100 times
  update_every <- max(1L, floor(n / 100))
  
  shiny::withProgress(message = "Simulating tournaments", value = 0, {
    for (i in seq_len(n)) {
      # Simulate the group stage
      qualifiers <- qualify_r32()
      
      # Simulate the knockout stage
      ko <- play_knockout(qualifiers = qualifiers, calendar = calendar_ko)
      
      # Identify the teams that reached each stage
      reached <- list(
        R32 = qualifiers$team,
        R16 = ko$winner[ko$round == "R32"],
        QF = ko$winner[ko$round == "R16"],
        SF = ko$winner[ko$round == "Quarter"],
        Final = ko$winner[ko$round == "Semi"],
        Champion = ko$winner[ko$round == "Final"]
      )
      
      # Check the number of teams at each stage
      if (!identical(unname(lengths(reached)), c(32L, 16L, 8L, 4L, 2L, 1L))) {
        stop("Unexpected number of teams at a tournament stage.")
      }
      
      # Update the counters
      for (stage in stages) {
        participants <- reached[[stage]]
        
        if (anyNA(participants) ||
            anyDuplicated(participants) ||
            !all(participants %in% rownames(counts))) {
          stop("Invalid participants at stage: ", stage)
        }
        
        counts[participants, stage] <-
          counts[participants, stage] + 1
      }
      
      # Display calculation progress
      if (i %% update_every == 0L || i == n) {
        shiny::setProgress(value = i / n,
                           detail = paste(i, "of", n, "tournaments"))
      }
    }
  })
  
  # Convert counts to percentages
  result <- data.frame(
    team = rownames(counts),
    elo = unname(rating[rownames(counts)]),
    100 * counts / n,
    row.names = NULL,
    stringsAsFactors = FALSE
  )
  
  # Sort by title probability, then final probability
  result <- result[order(-result$Champion, -result$Final), ]
  
  rownames(result) <- NULL
  
  result
}
```

```{r server, echo=FALSE}
# Define the tournament stages
stage_columns <- c("R32", "R16", "QF", "SF", "Final", "Champion")

stage_labels <- c(
  "Round of 32",
  "Round of 16",
  "Quarter-finals",
  "Semi-finals",
  "Final",
  "Champion"
)

# Run Monte Carlo when the button is clicked
mc_run <- eventReactive(input$run_mc, {
  
  n <- input$n_simulations
  seed <- input$simulation_seed
  
  validate(
    need(
      length(n) == 1L && !is.na(n) &&
        n >= 100 && n <= 50000 && n == floor(n),
      "Choose an integer between 100 and 50,000."
    ),
    need(
      length(seed) == 1L && !is.na(seed) &&
        seed >= 1 && seed <= 1000000 && seed == floor(seed),
      "Choose an integer seed between 1 and 1,000,000."
    )
  )
  
  set.seed(as.integer(seed))
  
  list(
    predictions = run_mc_shiny(as.integer(n)),
    n = n,
    seed = seed
  )
})

# Display simulation status
output$simulation_status <- renderText({
  
  if (input$run_mc == 0) {
    return("Click 'Run simulations' to generate predictions.")
  }
  
  run <- mc_run()
  
  paste(
    format(run$n, big.mark = ","),
    "simulations completed — seed:",
    run$seed
  )
})


# TAB 1: display the ten favourites
output$favourites_chart <- renderPlotly({
  
  req(input$run_mc > 0)
  
  # Select the ten teams with the highest title probabilities
  top <- head(mc_run()$predictions, 10)
  
  # Order teams for the horizontal bar chart
  top$team <- factor(
    top$team,
    levels = rev(top$team)
  )
  
  plot_ly(
    top,
    x = ~Champion,
    y = ~team,
    type = "bar",
    orientation = "h",
    marker = list(color = "#2463A0"),
    hovertemplate =
      "%{y}<br>Title probability: %{x:.1f}%<extra></extra>"
  ) %>%
    layout(
      title = "Who is most likely to win?",
      
      xaxis = list(
        title = "Probability of winning",
        ticksuffix = "%",
        rangemode = "tozero"
      ),
      
      yaxis = list(
        title = "",
        automargin = TRUE
      )
    )
})


# TAB 2: display the selected team's name
output$team_title <- renderText({
  
  input$selected_team
})


# Display the selected team's group and Elo rating
output$team_details <- renderText({
  
  req(input$selected_team)
  
  team_id <- names(teams)[
    match(input$selected_team, unname(teams))
  ]
  
  paste0(
    "Group: ",
    substr(team_id, 1, 1),
    " | Elo: ",
    unname(rating[input$selected_team])
  )
})


# Display the selected team's probability of reaching each stage
output$team_chart <- renderPlotly({
  
  req(input$run_mc > 0)
  
  predictions <- mc_run()$predictions
  
  # Select the results for the chosen team
  selected <- predictions[
    predictions$team == input$selected_team,
    ,
    drop = FALSE
  ]
  
  # Extract its probabilities for each tournament stage
  probabilities <- as.numeric(
    unlist(
      selected[stage_columns],
      use.names = FALSE
    )
  )
  
  chart_data <- data.frame(
    stage = factor(
      stage_labels,
      levels = stage_labels
    ),
    probability = probabilities
  )
  
  plot_ly(
    chart_data,
    x = ~stage,
    y = ~probability,
    type = "bar",
    marker = list(color = "#008577"),
    hovertemplate =
      "%{x}<br>Probability: %{y:.1f}%<extra></extra>"
  ) %>%
    layout(
      title = "Probability of reaching each stage",
      
      xaxis = list(
        title = "",
        automargin = TRUE
      ),
      
      yaxis = list(
        title = "Probability",
        range = c(0, 100),
        ticksuffix = "%"
      )
    )
})
```

## Results

We simulated the World Cup 1000 times to explore three main questions : which teams are the favourites to win the title, how far each team is expected to progress and which group-stage outcomes were the most surprising compared with the model's predictions.

```{r run_mc, cache=TRUE}
#fix the seed
set.seed(123)

# run simulation
mc_results <- run_cup_mc(1000)
```

### Who are the favourite ?

```{r favourite, fig.align='center'}
# select top 10 
top_10 <- mc_results %>%
  arrange(desc(Champion), desc(Final)) %>%
  slice_head(n = 10) %>%
  mutate(team = factor(team, levels = rev(team)))


# Bar chart
champion_plot <- plot_ly(
  top_10,
  x = top_10$Champion,
  y = top_10$team,
  type = 'bar',
  orientation = 'h'
) %>%
  layout(title = "Who is most likely to win the World Cup ?",
         xaxis = list(title = "Probability of winning", ticksuffix = "%"))

# show
htmltools::div(
  style = "display: flex; justify-content: center;",
  champion_plot
)
```

```{r values_champions, include=FALSE}
first_team <- as.character(top_10$team[1])
first_prob <- top_10$Champion[1]

second_team <- as.character(top_10$team[2])
third_team <- as.character(top_10$team[3])

top_three_prob <- sum(top_10$Champion[1:3])

last_team <- as.character(top_10$team[10])
last_prob <- top_10$Champion[10]

prob_ratio <- first_prob / last_prob
```

Across the 1000 simulations, `r first_team` has the highest estimated probability of winning the World Cup at `r round(first_prob,1)`% , followed by `r second_team` and `r third_team`. Together, these three teams account for `r round(top_three_prob,1)`% of the title probability, which suggests that the model concentrates much of the winning potential among them.

`r last_team` ranks tenth with a probability of `r last_prob`%. `r first_team` is therefore approximately `r round(prob_ratio,1)` times as likely to win as `r last_team`, which highlights a significant variation in estimated title probabilities even among the ten leading teams.

### How far are teams expected to go ?

Although `r first_team`, `r second_team` and `r third_team` account for most of the estimated title probability, it can also be useful to examine how far each team is likely to progress in the tournament according to the team's strength.

```{r stage_heatmap, fig.width=10, fig.height=15, fig.align='center'}
# order of the stages
stages <- c("R32", "R16", "QF", "SF", "Final", "Champion")

# sort team by Elo
team_order <- mc_results %>%
  arrange(desc(elo)) %>%
  pull(team)

# transform data
heatmap_data <- mc_results %>%
  pivot_longer(cols = all_of(stages),
               names_to = "stage",
               values_to = "probability") %>%
  mutate(
    stage = factor(stage, levels = stages),
    team = factor(team, levels = rev(team_order)),
    
    # avoid display 0%
    label = if_else(probability == 0, "<0.1", sprintf("%.1f", probability))
  )

# create heatmap
heatmap_plot <- plot_ly(
  data = heatmap_data,
  x = ~ stage,
  y = ~ team,
  z = ~ probability,
  type = "heatmap",
  colors = c("#F3F7FC", "#084594"),
  zmin = 0,
  zmax = 100,
  text = ~ paste0(label),
  texttemplate = "%{text}",
  textfont = list(size = 10),
  
  hovertemplate = paste0(
    "<b>%{y}</b><br>",
    "%{x}<br>",
    "Probability: %{z:.1f}",
    "<extra></extra>"
  ),
  colorbar = list(title = "Probability", ticksuffix = "%"),
  width = 870,
height = 1100) %>%
  layout(
    title = "How far could each team go?",
    xaxis = list(title = "", side = "top"),
    yaxis = list(title = "")
  )

# show
htmltools::div(
  style = "
    max-height: 600px;
    overflow-y: auto;
    display: flex;
    justify-content: center;
  ",
  heatmap_plot
)
```

The heatmap displays each team's probability of reaching each stage, with teams ordered by Elo rating from highest to lowest. It shows that a higher ranking does not always correspond to a higher probability of advancing. Indeed, as seen for the Round of 32, advancement probabilities would be expected to generally decrease as Elo ratings decline. However, this pattern is not strictly observed, which suggests that tournament progression is influenced not only by team strength.

### Which group-stage outcomes were the most surprising ?

We compare each team's actual group-stage outcome with its simulated probability of reaching the round of 32. This allows us to identify unexpected qualifications, where teams advanced despite a low predicted probability, and unexpected eliminations, where teams failed to qualify despite a high predicted probability.

The actual results were :

```{r actual_results, warning=FALSE}
# load data
actual_results <- read.csv2(
  here("data", "actual_stage.csv"),
  colClasses = "character",
  stringsAsFactors = FALSE,
  fileEncoding = "UTF-8"
) %>%
  rename(team = country) %>%
  mutate(team = trimws(team), actual_stage = toupper(trimws(actual_stage)))

# show
actual_results_table <- actual_results %>%
  gt() %>%
  
  tab_header(
    title = "Actual Tournament Results"
  ) %>%
  cols_label(
    team = "Team",
    actual_stage = "Stage reached"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = cell_borders(
      sides = c("left", "right"),
      color = "grey80",
      weight = px(1)
    ),
    locations = cells_body(
      columns = everything()
    )
  ) %>%
  tab_style(
    style = cell_borders(
      sides = c("left", "right"),
      color = "grey70",
      weight = px(1)
    ),
    locations = cells_column_labels(
      columns = everything()
    )
  )

htmltools::div(
  style = "max-height: 500px; overflow-y: auto;",
  actual_results_table
)
```

The qualification probability is the proportion of Monte Carlo simulations in which a team reached the round of 32. We compare this probability with what actually happened : if a team actually qualified we use the qualification probability. If it was eliminated, we use $1 - p$, where $p$ is the qualification probability. A lower probability for the actual outcome indicates a more surprising result according to the simulations.

```{r group_surprises_plot, warning=FALSE}
# group data
group_surprises <- mc_results %>%
  select(team, R32) %>%
  left_join(actual_results, by = "team") %>%
  mutate(
    qualified = actual_stage != "GROUP",
    outcome = if_else(qualified, "Qualified", "Eliminated"),
    
    # observed probability
    observed_probability = if_else(
      qualified,
      R32,
      100 - R32
    )
  ) %>%
  arrange(R32, team) %>%
  mutate(position = row_number())


# plot
plot_height <- nrow(group_surprises) * 25

surprises_plot <- plot_ly(
  data = group_surprises,
  x = ~R32,
  y = ~position,
  color = ~outcome,
  colors = c(
    "Qualified" = "#008577",
    "Eliminated" = "#C45A35"
  ),
  type = "scatter",
  mode = "markers",
  marker = list(size = 10),
  text = ~team,
  height = plot_height,
  width = 850,
  hovertemplate = paste0(
    "<b>%{text}</b><br>",
    "Probability: %{x:.1f}%",
    "<extra></extra>"
  )
) %>%
  layout(
    title = "Group-stage surprises",
    
    xaxis = list(
      title = "Simulated probability of reaching the Round of 32",
      range = c(0, 100),
      tickvals = seq(0, 100, 25),
      ticksuffix = "%"
    ),
    
    yaxis = list(
      title = "",
      tickmode = "array",
      tickvals = group_surprises$position,
      ticktext = group_surprises$team
    ),
    
    legend = list(
      title = list(text = "Actual outcome")
    )
  )

# show
htmltools::div(
  style = "
    max-height: 650px;
    overflow-y: auto;
    display: flex;
    justify-content: center;
  ",
  surprises_plot
)
```

The teams are ordered by their predicted probability of reaching the round of 32. The most unexpected outcomes are qualifications, shown in green, at the low-probability end, and eliminations, shown in orange, at the high-probability end. The ten most surprising outcomes are listed below :

```{r top_surprises, warning=FALSE}
# group data
top_surprises <- group_surprises %>%
  arrange(observed_probability) %>%
  slice_head(n = 10) %>%
  select(team,
         qualification_probability = R32,
         outcome,
         observed_probability)

# show
top_surprises %>%
  gt() %>%
  cols_label(
    team = "Team",
    qualification_probability = "Qualification probability",
    outcome = "Actual outcome",
    observed_probability = "Observed outcome probability"
  ) %>%
  fmt_number(
    columns = c(
      qualification_probability,
      observed_probability
    ),
    decimals = 1,
    suffixing = FALSE
  ) %>%
  cols_align(
    align = "center",
    columns = everything()
  )
```

The top ten surprises are selected as the outcomes with the lowest observed-outcome probabilities according to the model.

The most surprising group-stage outcome is Uruguay's elimination, as the team had a high qualification probability but was actually eliminated. On the other hand, Ghana had a low qualification probability but still qualified.

## Conclusion

We reconstructed the 2026 World Cup using the official group composition and knockout schedule, with match outcomes simulated using Elo ratings and a separate model for draws. By repeating the tournament 1000 times, we estimated each team's probability of reaching each stage and winning the title.

The heatmap showed that team strength alone does not determine tournament outcomes. Some lower-rated teams had higher qualification probabilities than stronger teams. In addition, comparing simulated qualification probabilities with actual outcomes also revealed several unexpected results. Some teams were eliminated despite having a high probability of qualifying, while others qualified despite having a relatively low probability of doing so.
