---
title: "world_cup_prediction"
output: 
  html_document
runtime: shiny
---

Building on the previous analyses, this section moves from describing and grouping team characteristics to testing their predictive value. We investigate whether pre-tournament information can help distinguish teams that progress to at least the Round of 16 in the 2026 World Cup.

## Prepare prediction dataset

```{r}
# Load required packages
library(tidyverse)
library(here)
library(lubridate)
library(ranger)
library(yardstick)
library(rsample)
library(ggrepel)

# Load data
d.full <- read.csv(here("data", "elo_ratings.csv"))
d.wc <- read.csv(here("data", "wc_elo_ratings.csv"))

# Extract the pre-tournament snapshot
# Chose2025-12-31 as the last annual snapshot before the start of the 2026 World Cup
d.pre2026 <- d.full |>
  filter(
    year == 2025,
    snapshot_date == as.Date("2025-12-31") 
  ) 

# Extract the 2026 tournament outcomes
d.outcome2026 <- d.wc |>
  select(
    country,
    wc2026_exit_round
  )

# Combine pre-tournament team information from 2025 with the observed 2026 World Cup outcome.
# ・Predictors = information available before the 2026 World Cup
# ・Outcome = actual 2026 World Cup result
d.predict2026 <- d.pre2026 |>
  select(
    country,
    rating,
    rank,
    rating_max,
    wins,
    matches_total,
    is_host,
    confederation
  ) |>
  left_join(
    d.outcome2026,
    by = "country"
  )

# Check that all 48 teams are included
nrow(d.predict2026)

# Check for unmatched teams
sum(is.na(d.predict2026$wc2026_exit_round))
```

The resulting dataset contains 48 teams, corresponding to the 48 teams participating in the 2026 World Cup. No teams have missing tournament outcomes after the join, confirming that all 48 teams were successfully matched with their observed 2026 results.

## Define the prediction outcome

```{r}
d.full |>
  filter(year == 2026) |>
  count(snapshot_date)
```

```{r}
# advanced = 1 indicates that a team reached the Round of 16
# advanced = 0 indicates elimination in the group stage or Round of 32

d.predict2026 <- d.predict2026 |>
  mutate(
    advanced = if_else(
      wc2026_exit_round %in% c("R16", "active"),  # active:if still in, else the round eliminated
      1,
      0
    )
  )

# Outcome distribution
table(d.predict2026$advanced)
```

The outcome distribution contains 32 teams classified as not advancing (advanced = 0) and 16 teams classified as advancing to at least the Round of 16 or remaining active at the 2026-07-07 snapshot (advanced = 1). Thus, one third of the teams had either reached the Round of 16 or were still active at the time of the 2026-07-07 snapshot.

The prediction problem under investigation is therefore formulated as follows: To what extent can pre-tournament team characteristics distinguish between teams that progress to at least the Round of 16 and those eliminated in the earlier stages of the tournament?

## Feature engineering

```{r}
# create two new variables:
# ・rating_gap: measures how far a team's current Elo rating is below its historical maximum Elo rating. 
# 　A larger value therefore indicates a larger distance from the team's historical peak.
# ・win_rate: represents the proportion of recorded matches that the team has won.
# ・is_host: 1 if the team is hosting the tournament (USA, Canada, Mexico), else 0.
# ・confederation: UEFA / CONMEBOL / CONCACAF / CAF / AFC / OFC

d.predict2026 <- d.predict2026 |>
  mutate(
    rating_gap = rating_max - rating,
    win_rate = wins / matches_total,
    is_host = factor(is_host),  
    confederation = factor(confederation) 
  ) # Convert into factors, to make them easier to use in the statistical models that follow

d.predict2026 |>
  select(country, rating_gap, win_rate)
```

The resulting dataset contains the original team-level predictors together with the newly created `rating_gap` and `win_rate` variables. `is_host` and `confederation` have also been converted to factors for use in the statistical models.

## Select predictors

```{r}
# Use a correlation matrix to check the correlations between the variables.
d.predict2026 |>
  select(
    rating,
    rank,
    rating_gap,
    win_rate
  ) |>
  cor(use = "complete.obs")
```

The correlation matrix shows a very strong negative correlation between `rating` and `rank` (r = -0.971). This indicates substantial redundancy between the two predictors and raises a clear concern about multicollinearity. Therefore, `rank` was excluded, while `rating` was retained in the model.

The correlations among the remaining predictors are considerably weaker. Therefore, `rating_gap` and `win_rate` were retained as additional predictors.

## Fit the GLM Model

***GLM Model***

```{r}
# Fit a simplified logistic regression model
glm_model <- glm(
  advanced ~ rating + rating_gap + win_rate,
  data = d.predict2026,
  family = binomial
  # Because advanced only takes: 0 = did not make the Round of 16; 1 = made the Round of 16
  # So use:Logistic Regression / Binomial GLM
)

# Display the model results
summary(glm_model)
```

***GLM Model Interpretation***

- Elo rating is the only statistically significant predictor of reaching at least the Round of 16 (β = 0.010669, p = 0.00173). Higher Elo ratings are associated with higher odds of advancing, holding the other predictors constant.

- Rating gap (p = 0.77470) and win rate (p = 0.70809) are not statistically significant predictors in this model.

- The residual deviance (42.688) is lower than the null deviance (61.105), indicating that the model with predictors fits the observed outcomes better than an intercept-only model.

- The AIC is 50.688, which can be used to compare alternative models fitted to the same data; lower values generally indicate a better fit–complexity trade-off.

- The model converged normally after five Fisher scoring iterations, with no convergence warning reported.\

```{r}
# Calculate odds ratios
glm_results <- tibble(
  Predictor = names(coef(glm_model))[-1],
  Odds_Ratio = exp(coef(glm_model)[-1]),
  P_Value = summary(glm_model)$coefficients[-1, "Pr(>|z|)"]
)

glm_results
```

The odds ratios provide an interpretation of the estimated effects of the predictors on the likelihood of reaching at least the Round of 16.

- Rating: OR = 1.01. A one-point increase in Elo rating is associated with approximately a 1.1% increase in the odds of reaching at least the Round of 16, holding the other predictors constant. This association is statistically significant (p = 0.00173).

- Rating gap: OR = 1.00. The estimated effect is very small, and the association is not statistically significant (p = 0.775). Therefore, the rating gap does not provide clear evidence of an additional association with reaching the Round of 16 in this model.

- Win rate: OR = 0.121. The model estimates lower odds of reaching the Round of 16 at higher values of win rate. However, this association is not statistically significant (p = 0.708), so the estimated negative direction should not be interpreted as evidence of a meaningful relationship.\

## Generate final GLM probabilities for the 48 teams

```{r}
d.predict2026 <- d.predict2026 |>
  mutate(
    glm_probability = predict(
      glm_model,
      type = "response"
    )
  )

d.predict2026 |>
  select(
    country,
    rating,
    glm_probability,
    advanced
  ) |>
  arrange(desc(glm_probability))
```

The GLM was fitted using all 48 teams and the pre-tournament predictors. The resulting fitted probabilities represent the model's estimated probability that each team would reach at least the Round of 16. The teams are ordered from highest to lowest predicted probability. For example, Spain received the highest predicted probability (94%), followed by Argentina (90%) and France (85%).

## Visualize predicted probabilities

```{r}
# Prepare data for the probability plot
plot_data <- d.predict2026 |>
  arrange(glm_probability) |>
  mutate(
    position = row_number(),
    actual_result = factor(
      advanced,
      levels = c(0, 1),
      labels = c("Did not advance", "Advanced")
    ),
    prediction_error = if_else(
      (glm_probability >= 0.5 & advanced == 0) |
      (glm_probability < 0.5 & advanced == 1),
      "Error",
      "Correct"
    )
  )
```

```{r}
# Visualize predicted probabilities
ggplot(
  plot_data,
  aes(
    x = glm_probability,
    y = position
  )
) +
  geom_point(
    aes(
      fill = actual_result
    ),
    shape = 21,
    size = 4,
    color = "white"
  ) +
  geom_vline(
    xintercept = 0.5,
    linetype = "dashed"
  ) +
  geom_text(
    data = filter(plot_data, prediction_error == "Error"),
    aes(label = country),
    nudge_y = 0.5,
    size = 3
  ) +
  scale_x_continuous(
    limits = c(0, 1),
    labels = scales::percent
  ) +
  scale_y_continuous(
    breaks = NULL
  ) +
  labs(
    title = "Predicted Probability of Reaching the Round of 16",
    subtitle = "Labels show teams incorrectly classified by the GLM",
    x = "Predicted probability",
    y = NULL,
    fill = "Actual outcome"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.major.y = element_blank()
  )
```

The plot shows the GLM's fitted probability of reaching the Round of 16 for each of the 48 teams. The dashed vertical line at 50% represents the classification threshold: teams with fitted probabilities of at least 50% are classified as expected to advance, while those below 50% are classified as expected not to advance.

The fitted probabilities show a clear separation between many of the teams that advanced and those that did not. Most teams that did not advance (pink points) have fitted probabilities below 50%, while most teams that advanced (blue points) have probabilities above 50%. The fitted probabilities also vary substantially across teams, ranging from very low values for some non-advancing teams to above 90% for some teams with higher Elo ratings.

The plot also highlights several misclassified teams, whose country names are labelled. On the right side of the 50% threshold, teams such as Netherlands, Ecuador, Croatia, Germany, and Uruguay were assigned probabilities above 50% but did not actually advance. These are ***false positives***. Conversely, teams such as Belgium, Paraguay, Morocco, Mexico, Canada, the United States and Egypt advanced despite receiving probabilities below 50%, representing ***false negatives***.

This shows that the 50% classification rule does not correctly classify all 48 observed outcomes. Teams with fitted probabilities close to the 50% threshold are relatively uncertain cases under this classification rule, while teams at the extreme ends of the probability range receive more clearly differentiated fitted probabilities from the GLM.

## Conclusion

Within this sample, the GLM provided evidence of an association between the included pre-tournament predictors and reaching at least the Round of 16. Compared with the intercept-only model, the GLM showed a substantial reduction in deviance, indicating an improved overall model fit. Among the individual predictors included in the logistic regression, Elo rating was the only statistically significant predictor. Higher Elo ratings were associated with higher odds of advancing, while rating gap and win rate did not show statistically significant evidence of an association with tournament progression.

The GLM was fitted using the 48 teams and their pre-tournament characteristics. The resulting fitted probabilities provide a model-based summary of the estimated probability of reaching at least the Round of 16 for each team. When a 50% threshold was used for classification, the model produced both false-positive and false-negative classifications. Some teams received relatively high fitted probabilities but did not advance, while some teams with fitted probabilities below 50% nevertheless reached the Round of 16. This illustrates that the observed tournament outcomes are not completely separated by the pre-tournament predictors included in the model.

Overall, the results suggest that the GLM captured a meaningful relationship between pre-tournament team characteristics and tournament progression in this sample, while not perfectly classifying all observed outcomes. However, the findings should be interpreted with caution because the analysis is based on only 48 teams, the model was evaluated on the same data used for fitting, and other tournament-related factors not included in the model may also influence progression.
